File-based localization of numerical perturbations in data analysis pipelines

### 2020-02-28 paper.tex
Updates in methodo

@@

Figure~\ref{fig:overview-tool} shows an overview of our method. The output
of 2 computational conditions is compared through 3 main steps. At step 1,
we build a low-level provenance graph for the processing of each subject in
Condition 1, where nodes are OS processes created by the application and
edges are file dependencies. Such provenance graphs reconstruct a model of
the application workflow, regardless of the original programming language
and dependencies. They may, however, slightly differ across the processed
subjects since different execution paths might be followed due to
variations in the data. To address this issue, we cluster subjects based on
their provenance graph (step 2), resulting in a reduced set of $k$
representative provenance graphs. These $k$ graphs and their corresponding
subjects /TODO: ???/ are then used for process labeling (step 3), where
Condition 2 is executed step by step, comparing the files produced by each
process to their counterparts in Condition 1, making sure that temporary
files and files written by more than a single process are correctly saved.

@@

\subsection{Provenance capture and representation}

We use the \reprozip tool~\cite{rampin2016reprozip}
to capture: (1) the set of processes created by the
pipeline, using the \texttt{clone()} or \texttt{fork()} system call, and
(2) the set of files read, written and executed by each process, including
temporary files. \reprozip collects this information through the
\texttt{ptrace()} system call and stores it in a SQLite database.

\toolname reconstructs a \emph{process tree} starting from the first process
created by the pipeline and traversing processes through \texttt{clone()}
and \texttt{fork()} dependencies. It also creates a process \emph{graph} 
from this tree by adding edges corresponding to file dependencies between processes. A
file dependency is defined between processes A and B if a file written by A
is read by B. Figure~\ref{fig:simple_script} shows an example of a process
tree and graph constructed from the example pipeline in
Listing~\ref{listing:sample-script}.

@@

\begin{figure}
\centering
  \includegraphics[width=0.3\columnwidth]{images/simple_graph}
  \caption{Process tree and graph
  constructed from the example pipeline in
  Listing~\ref{listing:sample-script}.
  Processes that read or write
  temporary files are 
  represented with squares. Plain edges 
  represent the process tree (\texttt{fork()} or \texttt{clone()} 
  system calls). Dashed edges represent file dependencies: temporary 
  files are in purple and result files are in black.
  Every node in the tree is labeled using (1) a process id created by our
  reconstruction, (2) the name of the executable run by the process.
  Process 0 is the initial call to the \texttt{sh} interpreter, and
  processes 1, 3 and 4 are the calls to FSL bet, stats and maths made in
  Listing~\ref{listing:sample-script}. Process 2 is forked by process 1: it
  was captured by \reprozip while it did not appear in
  Listing~\ref{listing:sample-script}. 
}
  \label{fig:simple_script}
\end{figure}

\subsection{Subject clustering}

Different subjects may lead to different provenance graphs, due to
variations that may be coming from differences in data types or
cardinality. For instance, some images may have been acquired multiple
times in some subjects, for noise-reduction purposes. Some of these
differences can be neglected, for instance, when a data decompression step
is present at the beginning of the execution for some subjects only, but
other differences cannot be ignored, for instance when complete different
processing paths are used in different subjects.

To reduce the number of provenance graphs to be examined, we cluster
process trees using agglomerative hierarchical clustering, as implemented
in SciPy~\cite{oliphant2007scipy}. We use the tree edit
distance~\cite{zhang1989simple} between process trees, as implemented in
the zss Python package. This distance is defined as the minimum number of
edit operations to transform one tree into the other. Three edit operations
are considered: node label modification, node removal, and node insertion.
Each operation has an associated cost of 1. 
//If threshold is 0, why do we even need the edit distance?//

\subsection{Graph analysis}

We clustered subjects into different types that each one refers to a specific graph.
The processes of the graph of each subject are labeled into two categories. First,
processes that read files that do not have differences and write files that
do not have differences are labeled \emph{transparent}
(Figure~\ref{fig:green}). Second, processes that read files that do not
have differences but write files that have differences \emph{create}
differences (Figure~\ref{fig:red}). In addition, processes that read files
that have differences and write files that also have differences are called
\emph{undetermined} (Figure~\ref{fig:grey}). Processes are also labeled
undetermined when they read or write temporary files, or produce multi-write
files. To resolve undetermined processes, hooks must be added to the pipeline to 
execute these processes on files without differences, and to capture
intermediary file states.

\begin{figure}%\centering
\centering
    \begin{subfigure}{0.3\linewidth}
      \centering
        \includegraphics[scale=0.36]{images/green.png}
        \caption{Transparent}
        \label{fig:green}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.35\linewidth}
      \centering
    \includegraphics[scale=0.36]{images/red.png}
    \caption{Creates differences}
    \label{fig:red}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.3\linewidth}
      \centering
    \includegraphics[scale=0.36]{images/grey.png}
    \caption{Undetermined}
    \label{fig:grey}
\end{subfigure}
    \caption{Different type of labeled process based on the input/output files.
  Dashed edges refer to the file dependencies between processes A and B 
  if a file written by A is read by B.}
    \label{fig:processes}
\end{figure}

\subsubsection{Capturing temporary files}

The label of processes that read or 
write temporary files is undetermined because these files are deleted during 
the execution. To address this issue, we replace in Condition 1 every process P that 
writes temporary files by a wrapper that first calls P and then
saves all its output files to a read-only directory. This replacement is
done by modifying the PATH environment variable to point to a directory
containing a modified version of the executable called by the process. The
modified executable will run the original executable and then save its
output files (see Listing~\ref{listing:modify-script}). 
This 
solution does not cover the temporary files that are removed by P 
itself, but this is not a problem since these files do not play any role in 
the subsequent steps of the pipeline.

@@

  \end{minted}
    \caption{Modified version of the script in Listing~\ref{listing:sample-script} to save intermediary files.}
    \label{listing:modify-script}
  \end{listing}

\subsubsection{Capturing multi-write files}

Files written by multiple processes also lead 
to undetermined labels. For a file F 
written by all processes in \textbf{P} = \{$P_{1}$, \ldots $P_{n}$\}, we 
(1) check that processes in \textbf{P} do not write concurrently to F, 
(2) establish an order on \textbf{P} based on the creation timestamp 
of the processes, (3) replace every process $P_{i}$ in \textbf{P} by 
a wrapper that first calls $P_{i}$ and then saves F to a read-only 
directory. Thus, multiple versions of F are saved and used in the 
analysis.

Furthermore, cycles may be present in the process graph in case a file 
was written by more than one process. We remove such cycles by removing 
file edges between processes A and B when A's process creation 
timestamp is posterior to B's or when A=B. Indeed, such edges cannot 
happen in practice unless A and B were running concurrently, which we 
assume is not the case.

As an example, in the sample pipeline in Listing~\ref{listing:sample-script}, process \texttt{2\#bet2} writes a temporary file.
Also, file \emph{T1w\_brain\_bin.nii.gz} is first written by \texttt{2\#bet2} and then overwritten by process \texttt{3\#fslmaths}.
The complete list of the processes that write intermediary files is captured in 
Listing~\ref{listing:json_process}.
Using this list, we modify processes in Condition 1 and then re-execute pipeline 
to save them in a backup folder.

@@

\subsubsection{Labelling undetermined processes}

When all the result files are captured in Condition 1, the tool 
is ready to label the pipeline processes in Condition 2.
However, another challenge is to label undetermined processes that read files with 
differences and write files with differences too. 
In this case, it is not possible to determine from the pipeline results 
whether the process creates differences or 
is transparent.
To address this issue, we set hooks in the pipeline 
in Condition 2. Hooks are set by replacing the pipeline 
processes with a wrapper script.
The replacement is done by modifying the PATH environment variable, as explained before. 
The wrapper script labels each process 
as \emph{transparent} or \emph{create differences} after running 
the original executable. 
Then, it copies the results obtained from the same process in 
Condition 1 to the pipeline output in Condition 2 if 
it is invoked with the arguments of a process that created differences.
Differences are identified by comparing the checksum of the output files for 
the same process in both conditions.
Besides, to measure the magnitude of the differences, we can add distance functions specific to file types.
We developed an incremental approach that consists of the following steps:

\begin{enumerate}
  \item Start the pipeline execution in Condition 2; 
        after hooking all the processes involved in the pipeline.
  \item Label each process as \emph{transparent} or \emph{create differences} 
        after running.
  \item Check if process is \emph{create differences} then: 
        copy the results produced by the same process in Condition 1 to the 
        pipeline output in Condition 2. 
  \item Continue steps 2 and 3 until the end of the pipeline execution.
\end{enumerate}

This algorithm creates a \emph{labeled process tree} at the end of pipeline execution 
which highlights processes that create differences.

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{images/iterative_modif}
  \caption{Illustration of the labelling approach.}
  \label{fig:iterations}
\end{figure}

Figure~\ref{fig:iterations} illustrates our incremental labelling 
process for the example in Figure~\ref{fig:simple_script}. At every 
step, processes that created differences are shown in red and other processes 
(transparent processes) are in green. Temporary files are not represented 
because they have been saved as previously described.

The four steps in Figure~\ref{fig:iterations} correspond to the 
steps of the labelling algorithm. 
At step one, process \texttt{1\#bet} is executed and then labeled 
as transparent (green) as it produced files without differences.
At step 2, \texttt{2\#bet2} 
is labeled as difference creator (red) as it produced files with differences 
from files without differences. Therefore, the files produced by \texttt{2\#bet2} in  
Condition 2 are replaced with the files produced by \texttt{2\#bet2} in 
Condition 1.
At step 3, after running process \texttt{3\#fslstats}, it labeled as 
transparent as it writes files without differences.
At step 4, process \texttt{4\#fslmaths} is labeled as difference creator 
as the last process of the pipeline.
As a result of these 4 steps, the final process labelling is: 
\texttt{2\#bet2} and \texttt{4\#fslmath} are difference creators (red) 
and the other processes are transparent.

@@

@@

>>>>

\todo{explain the issue with propagating differences and with transient files}

@@

The steps implemented in \toolname to compare pipeline executions in
different conditions are outlined in Figure~\ref{fig:overview-tool}. First,
in Condition 1, a directed, rooted, bipartite provenance graph is built for
each subject, where nodes represent application processes or files, and
edges represent read or write file accesses (Step 1 in Figure
Figure~\ref{fig:overview-tool}). \emph{Transient} files are then identified as
 files that are either deleted during the execution, or modified by
multiple processes (Step 2). \toolname re-runs the pipeline to save these files
right after they are produced, resulting in a provenance \emph{DAG}
(Directed Acyclic Graph) in which file nodes have a single parent (in-degree
of 1) (Step 3). This DAG is then compared to the one produced from
Condition 2, in a step-by-step process that prevents the propagation of differences.
\toolname finally produces a labeled graph that identifies the
non-reproducible processes of the pipeline.

@@

  \todo{review and summarize assumptions:
  * No concurrent writes from multiple processes
  * A given process cannot modify the bytes it wrote to a file
  But we support non-concurrent writes by different processes. This is common in neuroimaging.
  }

\subsection{Recording of provenance graphs}

We use the \reprozip tool~\cite{rampin2016reprozip}
to capture: (1) the set of processes created by the
pipeline, using the \texttt{clone()} or \texttt{fork()} system call, and
(2) the set of files read, written and executed by each process, including
temporary files. \reprozip collects this information through the
\texttt{ptrace()} system call, with no required instrumentation of the pipeline.
\toolname then reconstructs a provenance graph by creating process and file
nodes from the \reprozip trace, and by adding directed edges corresponding
to file reads and file writes (Figure~\ref{fig:simple_script}).

\todo{Is every process linked to an executable?}

Provenance graphs may be subject-dependent, which justifies iterating
\toolname on multiple subjects. For instance, data acquisitions may be
repeated a different number of times in each subject, for noise-reduction
purposes. Some of these differences can be neglected, for instance when a
data decompression step is present at the beginning of the execution for
some subjects only, but other differences cannot be ignored, for instance
when complete different processing paths are used in different subjects.

@@

\begin{figure}
\centering
  \includegraphics[width=0.3\columnwidth]{images/simple_graph}
  \caption{Process graph
  constructed from the example pipeline in
  Listing~\ref{listing:sample-script}.
  Processes that read or write
  temporary files are 
  represented with squares. Plain edges 
  represent the process tree (\texttt{fork()} or \texttt{clone()} 
  system calls). Dashed edges represent file dependencies: temporary 
  files are in purple and result files are in black.
  Every node in the tree is labeled using (1) a process id created by our
  reconstruction, (2) the name of the executable run by the process.
  Process 0 is the initial call to the \texttt{sh} interpreter, and
  processes 1, 3 and 4 are the calls to FSL bet, stats and maths made in
  Listing~\ref{listing:sample-script}. Process 2 is forked by process 1: it
  was captured by \reprozip while it did not appear in
  Listing~\ref{listing:sample-script}. \todo{Tristan: check that, this is not a bipartite graph now}. 
}
  \label{fig:simple_script}
\end{figure}

\subsection{Capture of transient files}

\toolname captures temporary files in Condition 1 by replacing every
process P by a wrapper that first calls P and then saves the produced
temporary files to a read-only directory. This process replacement is done
without pipeline instrumentation, by pre-pending to the PATH environment
variable a directory containing a wrapper script named as the executable
called by the process (see Listing~\ref{listing:modify-script}).

% The temporary files that are removed by P 
% itself won't be saved, but this is not a problem since these files do not play any role in 
% the subsequent steps of the pipeline.

@@

  \end{minted}
    \caption{Wrapper script used to save temporary and modified files.}
    \label{listing:modify-script}
  \end{listing}

Files written by multiple processes are disambiguated as follows. For a
 file F written by the processes in \textbf{P} = \{$P_{1}$, \ldots,
 $P_{n}$\}, \toolname first checks that processes in \textbf{P} do not
 write concurrently to F, which would violate our assumptions. Then, it
 replaces every process $P_{i}$ by a \texttt{PATH}-based wrapper that first
 calls $P_{i}$ and then saves F to a read-only directory. In this way, the
 multiple versions of F are preserved for comparison. \toolname finally
 updates the provenance graph accordingly, so that all files in the graph
 have an in-degree of 1. This operation also makes the provenance graph
 acyclic, since we assumed that a process cannot modify a file once it was
 written \todo{check/prove that}.

As an example, in the sample pipeline in
Listing~\ref{listing:sample-script}, process \texttt{2\#bet2} writes a
temporary file. Also, file \emph{T1w\_brain\_bin.nii.gz} is first written
by \texttt{2\#bet2} and then overwritten by process \texttt{3\#fslmaths}
\todo{it's not the case, check}. The complete list of the processes that
write intermediary files is captured in Listing~\ref{listing:json_process}.
Using this list, we modify processes in Condition 1 and then re-execute
pipeline to save them in a backup folder.

@@

\subsection{Process labelling}

Once all the files have been captured in Condition 1, including transient
files, \toolname labels pipeline processes through a step-by-step execution
in Condition 2. Each process in Condition 2 is run and its output files are
compared to the ones produced in Condition 1. If no differences are found,
the process is marked as reproducible. Otherwise, the process is marked as
irrreproducible and the output files produced in Condition 1 are copied to
Condition 2, to ensure that differences do not propagate further in the
pipeline. Processes are instrumented transparently, through a modification
of the \texttt{PATH} variable similar to the one described previously.

By default, differences in output files are identified by comparing file
checksums. Other comparison functions can also be defined for specific file
types, for instance to ignore sections of the file containing timestamps or
other execution-specific information.

Finally, \toolname creates a labeled provenance graph, 
highlighting non-reproducible processes.

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{images/iterative_modif}
  \caption{Illustration of the labelling approach. Non-reproducible processes are shown in red. \todo{update with bi-partite graph}}
  \label{fig:iterations}
\end{figure}

Figure~\ref{fig:iterations} illustrates our incremental labelling 
process for the example in Figure~\ref{fig:simple_script}. 
At step one, process \texttt{1\#bet} is executed and then labeled 
as reproducible (green) as it produced files without differences.
At step 2, \texttt{2\#bet2} 
is labeled as non-reproducible (red) as it produced files with differences 
from files without differences. Therefore, the files produced by \texttt{2\#bet2} in  
Condition 2 are replaced with the files produced by \texttt{2\#bet2} in 
Condition 1.
At step 3, \texttt{3\#fslstats} is labeled as 
reproducible as it writes files without differences.
At step 4, \texttt{4\#fslmaths} is labeled as non-reproducible 
as the last process of the pipeline.

@@

% \subsection{Subject clustering}

% To reduce the number of provenance graphs to be examined, we cluster
% process trees using agglomerative hierarchical clustering, as implemented
% in SciPy~\cite{oliphant2007scipy}. We use the tree edit
% distance~\cite{zhang1989simple} between process trees, as implemented in
% the zss Python package. This distance is defined as the minimum number of
% edit operations to transform one tree into the other. Three edit operations
% are considered: node label modification, node removal, and node insertion.
% Each operation has an associated cost of 1. 
% //If threshold is 0, why do we even need the edit distance?//

@@

Method:
* we assume no concurrent file accesses.

* we assume that the provenance graph is oriented, i.e., a process cannot read or write in the same file. Reasonnable in general but wouldn't work 
for, say, out of core computations.

Comment on the overhead, number of executions (3, 2 in condition 1, 1 in condition 2)

Applicability: no instrumentation required, fast.
But not able to capture dependencies in memory.
